---
title: 'Notebook3'
author: 'Alek Popovic, Firas Sada'
date: '2024-04-21'
output: html_document
---

# Predicting Top 100 Status from Player Stats

This notebook demonstrates various statistical methods to predict whether a player's other performance statistics qualify them for a Top 100 status, excluding the Wins Above Replacement (WAR) metric.

## Setup Libraries and Seed

Load necessary libraries for data manipulation and predictive modeling. Set a random seed for reproducibility.


```{r}

library(caret)
library(glmnet)
library(rpart)
library(randomForest)

set.seed(1122) # For reproducibility

```

## Data Loading and Initial Processing

Load the dataset and display the first few rows to understand the initial data structure.

```{r}

df = read.csv('dataset/processed_data.csv')
head(df)

```

## Data Preparation

Refine the dataset by removing the Wins Above Replacement column and rows where Top100Status is 'No'. Convert Top100Status to an integer for modeling purposes.


```{r}

# Remove the 'WinsAboveReplacement' column
df <- df[, !colnames(df) %in% 'WinsAboveReplacement']

# Remove rows where Top100Status is 'No'
df <- df[df$Top100Status != 'No', ]

# Convert the Top100Status column to integer
df$Top100Status <- as.integer(df$Top100Status)

# Shuffle the rows of 'df'
df <- df[sample(nrow(df)), ]

head(df)

```

## Train-Test Split

Create an 80-20 split of the data into training and testing sets to evaluate the performance of the models.

```{r}

# Create an 80-20 train-test split
index <- createDataPartition(df$Top100Status, p=0.8, list=FALSE)
df_train <- df[index, ]
df_test <- df[-index, ]

```

## Model Training and Evaluation Function

Define a function to fit models using Linear, Ridge, or Lasso Regression and calculate Mean Absolute Error (MAE) for performance evaluation.


```{r}

fit_and_evaluate <- function(model_formula, df_train, df_test, alpha_value = NULL) {
  
  if (is.null(alpha_value)) {
    # Fit Linear Regression
    model <- lm(model_formula, data = df_train)
    predictions <- predict(model, newdata = df_test)
  } else {
    # Fit Ridge or Lasso Regression
    model_matrix <- model.matrix(Top100Status ~ ., data = df_train)
    y <- df_train$Top100Status
    model <- glmnet(model_matrix, y, alpha = alpha_value)
    
    # Find the lambda that minimizes the cross-validation error
    cv_model <- cv.glmnet(model_matrix, y, alpha = alpha_value)
    lambda_optimal <- cv_model$lambda.min
    
    # Predict on test set
    test_matrix <- model.matrix(Top100Status ~ ., data = df_test)
    predictions <- predict(model, newx = test_matrix, s = lambda_optimal)
  }
  
  # Adjust prediction output for glmnet
  if (!is.null(alpha_value)) {
    predictions <- as.vector(predictions) # Convert matrix to vector if glmnet
  }
  
  # Calculate and return Mean Absolute Error
  mae <- mean(abs(predictions - df_test$Top100Status))
  return(mae)
  
}

```

## Evaluate Model Performance

Evaluate and display the Mean Absolute Error (MAE) for different regression models.

```{r}

# Linear Regression
linear_mae <- fit_and_evaluate(Top100Status ~ ., df_train, df_test)
cat('Linear Regression MAE:', linear_mae, '\n')

# Ridge Regression (alpha = 0 for ridge)
ridge_mae <- fit_and_evaluate(Top100Status ~ ., df_train, df_test, alpha_value = 0)
cat('Ridge Regression MAE:', ridge_mae, '\n')

# Lasso Regression (alpha = 1 for lasso)
lasso_mae <- fit_and_evaluate(Top100Status ~ ., df_train, df_test, alpha_value = 1)
cat('Lasso Regression MAE:', lasso_mae, '\n')

```

## Decision Tree Model

Train and evaluate a decision tree model to predict Top100Status.


```{r}

tree <- rpart(Top100Status ~ ., data = df_train, method='anova')

# Predict on the test set
predictions <- predict(tree, newdata=df_test)

# Calculate the Mean Absolute Error (MAE)
mae <- mean(abs(predictions - df_test$Top100Status))
cat('Decision Tree Regression MAE:', mae, '\n')

```

## Random Forest Model

Implement and assess a Random Forest model using a small number of trees.

```{r}

forest <- randomForest(Top100Status ~ ., data = df_train, ntree=5)

# Predict on the test set
predictions <- predict(forest, newdata=df_test)

# Calculate the Mean Absolute Error (MAE)
mae <- mean(abs(predictions - df_test$Top100Status))
cat('Random Forest Regression MAE:', mae, '\n')

```
